An upper bound to gas storage and delivery via pressure-swing adsorption in porous materials

### 2021-03-23 gas-adsorption/rsc-gas-ads-complete.tex
add cartoon from Cory
unique => only bound possible. dont need to justify novelty
improve intro

\includegraphics{head_foot/dates} & \noindent\normalsize{
    Both hydrogen
and natural gas are challenging to economically store onboard vehicles as
fuels, due to their low volumetric energy density at ambient conditions.
One strategy to densify
these gases is to pack the fuel tank with a porous
adsorbent material.
% 
The US Department of
Energy (DOE) has set volumetric deliverable capacity
targets which, if met, would help
enable commercial adoption of hydrogen/natural gas as transportation fuels.
%
%Many have attempted to establish theoretical upper bounds on the deliverable
%capacity via pressure-swing adsorption using simplified models of the
%gas-substrate and gas-gas interaction~\cite{gomez2014exploring,
%gomez2017impact, kaija2018high, lee2019predicting}; none have established a
%rigorous upper bound on the deliverable capacity.
%
Here, we present a
theoretical upper bound on the deliverable capacity of a gas in a rigid
porous
material via an isothermal pressure swing.
%
To provide an extremum, we consider
a substrate that provides a spatially uniform potential energy field for the
gas. Our bound is unique in that it relies directly on experimentally measured
properties of the pure gas.
%
We conclude
that the goals set by the DOE for room-temperature natural gas and hydrogen
storage are just barely theoretically possible.
The goals are likely to be impossible for any real,
rigid porous material because of steric
repulsion, which must reduce the capacity below our upper
bound. Limitations
to the scope of applicability of our upper bound may guide fuel tank design and
future material development. Firstly, one could avoid using an \emph{isothermal}
pressure swing by heating the adsorbent to drive off
trapped, residual gas. Secondly,
our upper bound assumes the material does not change its
structure in response to adsorbed gas, suggesting that flexible materials could
still satisfy the DOE targets.}

@@

%%%MAIN TEXT%%%%
\section{Introduction}
The transportation sector is dominantly powered by petroleum-based
fuels~\cite{davis2009transportation}. In the United States, it is responsible
for 36\% of energy-related carbon dioxide emissions \cite{useia} and $\sim$50,000
premature deaths per year associated with particulate matter and ozone
emissions~\cite{caiazzo2013air}. Natural gas and hydrogen (H$_2$) are
alternative transportation fuels that emit fewer/less pollutants than
petroleum-based fuels, and thus their widespread adoption could mitigate
climate change~\cite{mcglade2015geographical} and improve air quality for human
health. Further, as the finite global petroleum resources are declining
rapidly~\cite{sorrell2010global}, the development of technologies for the
widespread adoption of sustainable transportation fuels, such as hydrogen, is
critical.

@@

At ambient conditions, both methane and hydrogen gas possess a very low
volumetric energy density compared to gasoline. Consequently, under storage
space constraints in passenger vehicles, natural gas and hydrogen must be
densified for onboard storage to achieve a reasonable driving range on a
``full'' tank of fuel. Traditional densification approaches are liquefaction,
at cryogenic temperatures and atmospheric pressure, or compression, at room
temperature and high pressures. Both approaches require expensive
infrastructure at refilling stations and significant energy input; e.g., the
energy input to liquify hydrogen is $\sim$30\% of its energy
content~\cite{bossel2003energy}. Moreover, high-pressure storage tanks are
heavy, thick-walled, and non-conformable, while cryogenic storage tanks are
bulky, expensive, and afflicted by boil-off losses~\cite{hasan2009minimizing}.

A promising approach to densify natural
gas~\cite{makal2012methane,mason2014evaluating} and
hydrogen~\cite{suh2011hydrogen,garcia2018benchmark} for vehicular storage at
room temperature is through physical adsorption in nanoporous
materials~\cite{schoedel2016role}. The internal surfaces of porous materials
attract gas molecules through van der Waals, electrostatic, etc.\ interactions
to achieve a higher adsorbed gas density than the bulk gas at the same
temperature and pressure, allowing for room temperature and lower-pressure
storage and alleviating many drawbacks of high-pressure or low-temperature
storage.

@@

For a vehicle employing a porous material to store natural gas or hydrogen, the
(volumetric) \emph{deliverable capacity} of the gas is the primary
thermodynamic property that determines the driving
range~\cite{mason2014evaluating}. The adsorbed gas storage tank delivers the
gaseous fuel to the engine via an (assumed) isothermal pressure
swing~\cite{sircar2002pressure}. The deliverable capacity (see
Fig.~\ref{fig:delcap}) is the density of the gas in the material at the storage
pressure $\pfull$ minus the residual gas that remains adsorbed at the lowest
pressure $\pempty$ such that sufficient flow is maintained to feed the engine.
For commercial feasibility, the US Department of Energy (DOE) has set
deliverable capacity targets for both adsorbed methane and hydrogen storage and
delivery for vehicles. For methane, the Advanced Research Projects
Agency--Energy (ARPA-E) set a deliverable capacity target of 315\ L STP CH$_4$
per L of adsorbent at 298 K using a 65\ bar to 5.8\ bar pressure
swing~\cite{simon2015materials}. For hydrogen, the DOE set a series of
progressive targets at five year intervals, with the ultimate target of 0.05 kg
H$_2$/L~\cite{h2targetsDOE} using a 100\ bar to 5\ bar pressure swing at a
minimum of -40 $^\circ$C~\cite{allendorf2018assessment}. Thus far, despite the
emergence of highly tunable materials with large surface areas, such as
metal-organic frameworks~\cite{furukawa2013chemistry} and covalent organic 
frameworks~\cite{}, no porous material has met these deliverable capacity targets.

@@

To set realistic performance targets and optimally allocate research resources,
in this work, we present a theoretical framework that places an intrinsic upper
limit on the deliverable capacity of any gas in a rigid porous material and
uses as input the experimentally measured properties of the bulk gas. Our
extremum is provided by a substrate that offers a spatially uniform potential
energy field felt by the gas. Applying our framework to methane and hydrogen
gas, we find the US DOE deliverable capacity targets for natural gas and
hydrogen storage and delivery are theoretically possible, but sufficiently
close to the upper bound as to be impractical for any real, rigid porous
material. Optimistically, new paradigms outside the scope of applicability of
our theoretical framework, such as gas-induced structural transitions of the
material, hold promise for meeting these targets, as evidenced by flexible MOF
Co(bdp) which currently boasts the largest methane deliverable
capacity~\cite{mason2015methane}.

@@

\cbstart
\section{Flexible two-phase model}
\begin{figure}
    \centering
    \includegraphics[width=0.95\columnwidth]{methane-298-flexible}
    \caption{Deliverable capacity of methane in a flexible substrate at 298\ K as a function of the free energy difference 
      between open and closed phases. The experimental deliverable capacity for Co(bdp) is shown for
       comparison~\cite{fixme}, along with the free energy difference between open and closed conformations~\cite{fixthis}.}
    \label{fig:methane-flexible}
\end{figure}
Although our limit does not directly apply to materials that change conformation in
response to gas adsorption, we can make an estimate of the benefits to be gained
materials such as \davidsays{FIXME AND CITE} which undergo a phase change in the
presence of gas.  We can model such a material as a two-phase system with one ``open''
phase and a ``closed'' phase in which all gas is expelled.  We assume there is a
free energy difference $\Delta G$ between the two phases in the absence of gas, with
the closed phase more stable.  In addition, we assume a $\V$ as above \davidsays{REFERENCE EQUATION?}
for the adsorption of gas into the open phase.  In order for the system to benefit
from the two phases, the open phase must be preferred at $\pfull$, with the closed phase
preferred at $\pempty$.  This requires that
\begin{align}
    V_{\text{open}}|\rho_{\text{open}}(\pfull)\V| &> |\Delta G| \\
    V_{\text{open}}|\rho_{\text{open}}(\pempty)\V| &< |\Delta G|
\end{align}
where $\rho_{\text{open}}(p)$ is the density of gas in the open phase of the material at
the given pressure.  This model has no upper bound on the deliverable capacity, since
$\V$ and $\Delta G$ can together be increased without bound.  We can, however, place a
bound for a fixed $\Delta G$, since if the gas is too strongly attracted to the open phase
the material will remain open at $\pempty$.  This upper bound is found by identifying the
attractive interaction $\V$ that solves the equation
\begin{align}
    \rho(\muempty{} + \V)\V = \Delta G/V_{\text{open}}
\end{align}
for $\V$, which gives the largest attraction consistent with the material closing
at the empty pressure.
Figure~\ref{fig:methane-flexible} illustrates
how this upper bound depends on $\Delta G$ for methane.  For comparison, we include the free energy
difference for Co(bdp)~\davidsays{CITE} and its measured deliverable capacity.  While this
upper bound is not rigorous, it suggests that a two-phase flexible material for methane storage
would benefit from significantly greater stability of the closed phase---provided the open phase
had correspondingly greater attraction for the gas molecules.  Figure~\ref{fig:hydrogen-flexible}
shows the corresponding result for hydrogen storage, indicating that for a flexible material to
reach the DOE ULTIMATE deliverable capacity would require far greater stability for the closed
phase, as well as an unprecedented level of attractive interaction for hydrogen.

>>>>

\includegraphics{head_foot/dates} & \noindent\normalsize{
Both hydrogen
and natural gas are challenging to economically store onboard vehicles as
fuels, due to their low volumetric energy density at ambient conditions.
One strategy to densify
these gases is to pack the fuel tank with a porous
adsorbent material.
% 
The US Department of
Energy (DOE) has set volumetric deliverable capacity
targets which, if met, would help
enable commercial adoption of hydrogen/natural gas as transportation fuels.
%
%Many have attempted to establish theoretical upper bounds on the deliverable
%capacity via pressure-swing adsorption using simplified models of the
%gas-substrate and gas-gas interaction~\cite{gomez2014exploring,
%gomez2017impact, kaija2018high, lee2019predicting}; none have established a
%rigorous upper bound on the deliverable capacity.
%
Here, we present a
theoretical upper bound on the deliverable capacity of a gas in a rigid
porous
material via an isothermal pressure swing.
%
To provide an extremum, we consider
a substrate that provides a spatially uniform potential energy field for the
gas. Our bound relies directly on experimentally measured
properties of the pure gas.
%
We conclude
that the deliverable capacity targets set by the DOE for room-temperature natural gas and hydrogen
storage are just barely theoretically possible.
The targets are likely to be impossible for any real,
rigid porous material because of steric
repulsion, which reduces the deliverable capacity below our upper
bound. Limitations
to the scope of applicability of our upper bound may guide fuel tank design and
future material development. Firstly, one could avoid using an \emph{isothermal}
pressure swing by heating the adsorbent to drive off
trapped, residual gas. Secondly,
our upper bound assumes the material does not change its
structure in response to adsorbed gas, suggesting that flexible materials could
still satisfy the DOE targets.}

@@

%%%MAIN TEXT%%%%
\section{Introduction}
The transportation sector is dominantly powered by petroleum-based
fuels~\cite{davis2009transportation}. 
Consequently, in the United States, it is responsible
for 36\% of energy-related carbon dioxide emissions \cite{useia} and $\sim$50,000
premature deaths per year associated with particulate matter and ozone
emissions~\cite{caiazzo2013air}. 
Natural gas and hydrogen (H$_2$) are
alternative transportation fuels that, if adopted widely,
could mitigate climate change~\cite{mcglade2015geographical} and improve air quality for human
health. 
More, the finite global petroleum resources are declining
rapidly~\cite{sorrell2010global}.
The development of technologies for the
widespread adoption of sustainable transportation fuels, such as hydrogen, is therefore
critical.

@@

At ambient conditions, both methane and hydrogen gas possess a very low
volumetric energy density compared to (liquid) gasoline. Consequently, under storage
space constraints in passenger vehicles, natural gas and hydrogen must be
densified for onboard storage to achieve a reasonable driving range on a
``full'' tank of fuel. Traditional densification approaches are liquefaction,
at cryogenic temperatures and atmospheric pressure, or compression, at room
temperature and high pressures. Both approaches require expensive
infrastructure at refilling stations and significant energy input; e.g., the
energy input to liquify hydrogen is $\sim$30\% of its energy
content~\cite{bossel2003energy}. Also, high-pressure storage tanks are
heavy, thick-walled, and non-conformable, while cryogenic storage tanks are
bulky, expensive, and afflicted by boil-off losses~\cite{hasan2009minimizing}.

A promising approach to densify natural
gas~\cite{makal2012methane,mason2014evaluating} and
hydrogen~\cite{suh2011hydrogen,garcia2018benchmark} for vehicular storage at
room temperature is through physical adsorption in nanoporous
materials~\cite{schoedel2016role}. The internal surfaces of porous materials
attract gas molecules through van der Waals, electrostatic, etc.\ interactions
to achieve a higher adsorbed gas density than the bulk gas at the same
temperature and pressure.
Nanoporous materials could thus allow for room temperature and lower-pressure
storage of natural gas and hydrogen and alleviate many drawbacks of high-pressure or low-temperature
storage.

@@

For a vehicle employing a porous material to store natural gas or hydrogen, the
(volumetric) \emph{deliverable capacity} of the gas is the primary
thermodynamic property that determines the driving
range on a given ``full'' tank of fuel~\cite{mason2014evaluating}. 
The adsorbed gas storage tank delivers the
gaseous fuel to the engine via an (assumed) isothermal pressure
swing~\cite{sircar2002pressure}. The deliverable capacity (see
Fig.~\ref{fig:delcap}) is the density of the gas in the material at the storage
pressure $\pfull$ minus the residual gas that remains adsorbed at the lowest
pressure $\pempty$ such that sufficient flow is maintained to feed the engine.
For commercial feasibility, the US Department of Energy (DOE) has set
deliverable capacity targets for both adsorbed methane and hydrogen storage and
delivery for vehicles. For methane, the Advanced Research Projects
Agency--Energy (ARPA-E) set a deliverable capacity target of 315\ L STP CH$_4$
per L of adsorbent at 298 K using a 65\,bar to 5.8\,bar pressure
swing~\cite{simon2015materials}. For hydrogen, the DOE set a series of
progressive targets at five year intervals, with the ultimate target of 0.05\,kg
H$_2$/L~\cite{h2targetsDOE} using a 100\ bar to 5\ bar pressure swing at a
minimum of -40 $^\circ$C~\cite{allendorf2018assessment}. Thus far, despite the
emergence of highly tunable materials with large surface areas, such as
metal-organic frameworks~\cite{furukawa2013chemistry} and covalent organic 
frameworks~\cite{diercks2017atom}, no porous material has met these deliverable capacity targets.

@@

To set realistic performance targets and optimally allocate resources to research efforts,
in this work, we present a theoretical framework that places an intrinsic upper
limit on the deliverable capacity of a pure gas in a rigid porous material and
uses as input the experimentally measured properties of the bulk gas. Our
extremum is provided by a substrate that offers a spatially uniform potential
energy field felt by the gas. Applying our framework to methane and hydrogen
gas, we find the US DOE deliverable capacity targets for natural gas and
hydrogen storage and delivery are theoretically possible, but sufficiently
close to the upper bound as to be impractical for any real, rigid porous
material. Optimistically, new paradigms outside the scope of applicability of
our theoretical framework, such as gas-induced structural transitions of the
material, hold promise for meeting these targets, as evidenced by flexible MOF
Co(bdp) which currently boasts the largest methane deliverable
capacity~\cite{mason2015methane}.

@@

\cbstart
\begin{figure}
    \centering
    \includegraphics[width=0.95\columnwidth]{figs/two-phase}
    \caption{Cartoon of our model.}
    \label{fig:flexible-cartoon}
\end{figure}
\section{Flexible two-phase model}
\begin{figure}
    \centering
    \includegraphics[width=0.95\columnwidth]{methane-298-flexible}
    \caption{Deliverable capacity of methane in a flexible substrate at 298\ K as a function of the free energy difference 
      between open and closed phases. The experimental deliverable capacity for Co(bdp) is shown for
       comparison~\cite{fixme}, along with the free energy difference between open and closed conformations~\cite{fixthis}.}
    \label{fig:methane-flexible}
\end{figure}
Although our limit does not directly apply to materials that change conformation in
response to gas adsorption, we can make an estimate of the benefits to be gained
materials such as \davidsays{FIXME AND CITE} which undergo a phase change in the
presence of gas.  We can model such a material as a two-phase system with one ``open''
phase and a ``closed'' phase in which all gas is expelled.  We assume there is a
free energy difference $\Delta G$ between the two phases in the absence of gas, with
the closed phase more stable.  In addition, we assume a $\V$ as above \davidsays{REFERENCE EQUATION?}
for the adsorption of gas into the open phase.  In order for the system to benefit
from the two phases, the open phase must be preferred at $\pfull$, with the closed phase
preferred at $\pempty$.  This requires that
\begin{align}
    V_{\text{open}}|\rho_{\text{open}}(\pfull)\V| &> |\Delta G| \\
    V_{\text{open}}|\rho_{\text{open}}(\pempty)\V| &< |\Delta G|
\end{align}
where $\rho_{\text{open}}(p)$ is the density of gas in the open phase of the material at
the given pressure.  This model has no upper bound on the deliverable capacity, since
$\V$ and $\Delta G$ can together be increased without bound.  We can, however, place a
bound for a fixed $\Delta G$, since if the gas is too strongly attracted to the open phase
the material will remain open at $\pempty$.  This upper bound is found by identifying the
attractive interaction $\V$ that solves the equation
\begin{align}
    \rho(\muempty{} + \V)\V = \Delta G/V_{\text{open}}
\end{align}
for $\V$, which gives the largest attraction consistent with the material closing
at the empty pressure.
Figure~\ref{fig:methane-flexible} illustrates
how this upper bound depends on $\Delta G$ for methane.  For comparison, we include the free energy
difference for Co(bdp)~\davidsays{CITE} and its measured deliverable capacity.  While this
upper bound is not rigorous, it suggests that a two-phase flexible material for methane storage
would benefit from significantly greater stability of the closed phase---provided the open phase
had correspondingly greater attraction for the gas molecules.  Figure~\ref{fig:hydrogen-flexible}
shows the corresponding result for hydrogen storage, indicating that for a flexible material to
reach the DOE ULTIMATE deliverable capacity would require far greater stability for the closed
phase, as well as an unprecedented level of attractive interaction for hydrogen.
